Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-09-15 manuscript/manuscript.tex
smoothed section 4

In addition to the matching values, we encountered a number of 
\textit{near-matches}, situations where the \texttt{cross\_street} 
and \texttt{intersection\_street} are nearly, but not exactly, identical. 
An example of such a near-match is when one field reads ``HARMON DR'' 
and the other reads ``HARMON RD.'' To identify these near-matches, 
we applied the \href{https://en.wikipedia.org/wiki/Hamming_distance}
{\textit{Hamming Distance}}, which measures the number of character 
changes required to make the two fields identical. For this analysis, 
we set the threshold for a near-match to a Hamming Distance of two (2).

@@

    \item \textbf{zip\_codes (computed field)}: This field has a 58\% error rate 
    and, as discussed in \ref{sec:case-study-zip-codes}, can lead to significant 
    analytical errors. We recommend deleting this field.
\end{itemize}

@@

\section{Recommendations} 
\label{sec:recommendations}
We propose the following recommendations by the Open Data Team to address
the data cleanliness issues addressed in this paper.

\begin{itemize}
	\item First, the Open Data Team will need 
	to undertake a dedicated effort to investigate these 
	data anomalies, confirm the veracity of the issue, 
	and engage with the appropriate Agency Open Data 
	Representatives to seek resolution. That effort will not 
	be easy or fast, and in some cases software changes may  
	be required. Coordination and mutual approval of any 
	changes will be key, requiring solid support 
	from senior leadership.

	\item Secondly, standards and protocols need to be 
	identified, agreed to, and commitment obtained. For example, 
	the domain of legal values, data accuracy, address normalization, 
	data usage, data definitions, etc..

	\item Many of the data integrity issues could potentially be 
	resolved through enforcement by software. For example, the 
	software should be able to easily prevent invalid values, such as zip 
	codes, and similarly for other data issues. One key area are is the 
	created\_date and closed\_date which are frequently cited 
	when determining responsiveness for various Agency services; software
	could conceivably prevent illogical situations such as an SR being
	``closed'' before it is ``opened''.

	\item The large spikes in SR closure and creation occurring at 
	midnight and noon almost certainly point to an automated 
	process from an external Agency integration. Such processes distort 
	the accurate capture of an SR duration and should be eliminated in
	favor of more accurate dates.

	\item As noted, there are six ``computed'' fields in the data export 
	that are not identified in the Data Dictionary. Several of 
	these computed fields suffer from significant data inaccuracies. We  
	recommend updating the Data Dictionary at a minimum and 
	possibly masking those fields from public usage until such time as 
	their accuracy can be ascertained.

	\item Removal of duplicate fields should be examined. In some cases, such 
	as the borough and park\_borough fields, the duplication is 100\%. In 
	other cases, the duplication is in the 80-90\% range. These near-duplicate 
	fields and 	their utility should be examined to determine if inclusion in the 
	311 SR dataset is warranted. Our analysis indicates that as much as 12\% 
	size reductions are achievable with a first pass, possibly more upon deeper
	analysis.

	\item Fields with extremely low data presence, such as taxi\_company\_borough 
	which is populated only 0.06\% of the time, should be evaluated 
	to determine if their inclusion is warranted.

	\item The Data Dictionary is dated March 2023. The issues addressed 
	in Section~\ref{sec:datadictionary} should be the bases for an update of
	the Data Dictionary.

	\item Excessive precision in the latitude, longitude, and location 
	fields is nonsensical.  We recommend no more than 
	four decimal places (0.000X), representing a distance of approximately 
	36.4 feet. Further decimal points strain credibility.

	\item The 11 address-related address fields (incident\_address, 
	street\_name, city, intersection\_street\_1 \& 2, cross\_street\_1 \& 2,  
	landmark, road\_ramp, bridge\_highway\_segment, and 
	taxi\_pick\_up\_location) should be standardized to USPS format. 
	Software packages can easily perform this standardization.

	\item There are multiple means by which missing data is 
	represented. These include ``UNKNOWN'' , ``N/A'', ``NA'', ``na'', and 
	a blank field, e.g. `` ''. Protocols should be established regarding 
	how to represent missing data.
\end{itemize}

>>>>

In addition to the matching values, we encountered a number of 
\textit{near-matches}, situations where the \texttt{cross\_street} 
and \texttt{intersection\_street} are nearly, but not exactly, identical. 
An example of such a near-match is when one field reads ``HARMON DR'' 
and the other reads ``HARMON RD.'' To identify these near-matches, 
we applied the \href{https://en.wikipedia.org/wiki/Hamming_distance}
{\textit{Hamming Distance}}, which measures the number of character
\jy{needs a quick definition and reference}
changes required to make the two fields identical. For this analysis, 
we set the threshold for a near-match to a Hamming Distance of two (2).

@@

  \item \textbf{zip\_codes (computed field)}:
    This field has a 58\% error rate and can lead to significant 
    analytical errors. We recommend deleting this field.
\end{itemize}

@@

\section{Recommendations} 
\label{sec:recommendations}
We propose the following recommendations for the Open Data Team to 
address the data cleanliness issues discussed in this paper.

\begin{itemize}
\item  Undertake a dedicated effort 
    to investigate these data anomalies, confirm the veracity of the 
    issues, and collaborate with the appropriate Agency Open Data 
    Representatives to find resolutions. This effort may be complex 
    and require software changes. Coordination and mutual approval 
    of changes will be key, requiring strong support from senior 
    leadership.

    \item Standards and protocols should be established, agreed upon, 
    and enforced. This includes defining the domain of legal values, 
    ensuring data accuracy, implementing address normalization, 
    clarifying data usage, and providing detailed data definitions.

    \item Resolve many potential data integrity issues
    through enforcement by software. For instance, software should 
    prevent invalid entries like erroneous zip codes. Similarly, fields 
    like \texttt{created\_date} and \texttt{closed\_date}, which 
    measure responsiveness, should have logic checks to prevent 
    illogical sequences, such as an SR being ``closed'' before 
    being ``opened.''

    \item The large spikes in SR creation and closure at midnight and 
    noon likely point to automated processes from external Agency 
    integrations. These distortions should be eliminated in favor of 
    capturing accurate SR durations with precise timestamps.

    \item As discussed, there are six ``computed'' fields that are not 
    listed in the Data Dictionary. Some of these fields contain 
    significant inaccuracies. We recommend updating the Data 
    Dictionary and, if necessary, masking these fields from public use 
    until their accuracy is improved.

    \item Duplicate fields should be reviewed for removal. In some 
    cases, such as \texttt{borough} and \texttt{park\_borough}, the 
    duplication is 100\%. In others, the match is 80-90\%. 
    These near-duplicates should be assessed to determine their 
    relevance in the 311 SR dataset. Initial analysis suggests that 
    removing these fields could reduce file size by up to 12\%, with 
    further reductions possible upon deeper review.

    \item Fields with very low data presence, such as 
    \texttt{taxi\_company\_borough}, populated only 0.06\% of the time, 
    should be evaluated to determine if they are necessary.

    \item The Data Dictionary, last updated in March 2023, should be 
    revised based on the issues discussed in Section~\ref{sec:datadictionary}.

    \item The precision of latitude, longitude, and location fields 
    should be reduced. We recommend limiting precision to four 
    decimal places (0.000X), which represents approximately 36.4 feet. 
    More decimal points strain credibility.

    \item The 11 address-related fields (\texttt{incident\_address}, 
    \texttt{street\_name}, \texttt{city}, \texttt{intersection\_street\_1 \& 2}, 
    \texttt{cross\_street\_1 \& 2}, \texttt{landmark}, \texttt{road\_ramp}, 
    \texttt{bridge\_highway\_segment}, and \texttt{taxi\_pick\_up\_location}) 
    should be standardized to USPS format. Software can easily 
    perform this standardization.

    \item There are multiple ways missing data is represented, such as 
    ``UNKNOWN'', ``N/A'', ``NA'', ``na'', and blanks (`` ''). Protocols 
    should be established to standardize the representation of missing data.
\end{itemize}
